DOpElib: Differential Equations and Optimization Environment; A Goal Oriented Software Library for Solving PDEs and Optimization Problems with PDEs

### 2012-12-10 Paper_May_2012/main.tex
Changed layout of the examples of tw and ww. There was a big gap between the subsections after the last changes.

%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
 \journalname{Computing}
%

@@

@@

\titlerunning{DOpElib}        % if too long for running head

\author{C. Goll \and T. Wick \and W. Wollner}

%\authorrunning{Short form of author list} % if too long for running head

\institute{Christian Goll \at
              University of Heidelberg,\\ Institute of Applied Mathematics, 
              \\Im Neuenheimer Feld 293/294, 69120 Heidelberg, Germany\\
              %Tel.: +49 622-154-6154\\
              \email{christian.goll@iwr.uni-heidelberg.de}           %  \\
           \and
Thomas Wick \at
              The Institute for Computational Engineering and Sciences,\\
The University of Texas at Austin,\\
Austin, Texas 78712, USA \\
              %Tel.: +1 512-672-7763\\
              \email{twick@ices.utexas.edu}   
 \and
Winnifried Wollner \at
              Universit\"at Hamburg\\
              Department of Mathematics\\
              Bundesstr. 55, 20146 Hamburg, Germany\\
              \email{winnifried.wollner@math.uni-hamburg.de}   
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In this article, we describe the  
\textit{Differential Equations and Optimization Environment} (\dope{}).
\dope{} is a software library that provides a unified interface 
to high level algorithms such as time-stepping methods, nonlinear solvers and 
optimization routines. This structure ensures 
that, first of all, the user is only required to write those sections of code 
that are specific to 
the considered problem. Second, the exchange of parts of the used routines is possible 
with only a few lines of code to change instead of large reimplementations.
The article illustrates the design principles and various features
of \dope{} and provides some 
numerical results as demonstration for the versatility of the software.

\keywords{software environment \and goal oriented interface design \and numerical solution of optimization problems and partial differential equations}
\subclass{49-04 \and  35-04}
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exists a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \citep{dealnew}, dune \citep{dune}, 
FEniCS \citep{fenics}
or commercial solvers as for instance ANSYS (with its Fluent package),
COMSOL, or simufact forming.
Finally, a lot of work is (still) done in Matlab.
In fact, these libraries provide very useful tools 
for a huge variety of applications.

@@

Now, we describe a program designed to compute a value 
$J(u)$, where $u=(v,p)$ is the solution of the Navier-Stokes 
equations \eqref{eq:ns} on the domain $\Omega$ as seen in 
Fig.~\ref{fig:example_ns} with  $f=0$, and $J$ is the drag-force 
of the enclosed circle. We assume given Dirichlet data for the velocity 
$v$ on the boundary $\Gamma_{D}=\Gamma_{in}\cup\Gamma_{wall}\cup\Gamma_{circ}$. 
This is the Benchmark-Case 2D-1 described in \cite{TuSchae96}.
\begin{figure}[h]
\centering
\resizebox{0.5\textwidth}{!}{\input{Pictures/example_ns}}
\caption{Flow around a cylinder with 
circle-center $C=(0.2,0.2)$ and radius $r=0.05$.}
\label{fig:example_ns}
\end{figure}

@@

\begin{figure}[hbt]
\centering
{\includegraphics[trim = 50mm 50mm 50mm 50mm, clip, width=\textwidth]{Pictures/local_grid_NSv2.png}}
\caption{Locally adapted grid after eight refinement steps.}
\label{fig:NS_local_grid}
\end{figure}

@@

The quantities of interest are evaluations of 
$x$- and $y$ displacement at the point $A(0) = (0.6,0.2)$
and the drag and lift forces acting on the cylinder and the elastic beam:
\begin{align}
\label{drag_lift_forces}
(F_D , F_L) 
= {\int_{S_f} \sigma_f \cdot n_f \, ds + 
\int_{\Gamma_i} \sigma_s \cdot n_s \, ds},
\end{align}
where $S_{f}$ denotes the path over the cylinder in the fluid part and
$\Gamma_i$ the interface between the elastic beam and the 
fluid.
% Bilder von FSI 2 
\begin{figure}[h]
\centering
{\includegraphics[width=4.5cm]{Pictures/visit_fsi_2_CNn_t_2e-2_global_3_biharmonic_mesh8070_scale.png}}
{\includegraphics[width=4.5cm]{Pictures/visit_fsi_2_CNn_t_2e-2_global_3_biharmonic_x_velo8070_scale.png}}
\caption{FSI 2 test case: mesh (left) and velocity profile in vertical 
direction (right) at time $t=16.14s$.}
\label{res:fsi_2_mesh_and_x_velo}
\end{figure}

@@

\begin{figure}[H]
\centering
{\includegraphics[width=4.5cm]{Pictures/ux_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=4.5cm]{Pictures/uy_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=4.5cm]{Pictures/Drag_fluid_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=4.5cm]{Pictures/Lift_fluid_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
\caption{FSI 2: The deflections of the beam, $u_x(A)$ and $u_y(A)$ (in $cm$), and 
the drag $F_D$ and the lift $F_L$ evaluation (in $kg/m\,s^2$) are displayed versus
time (in $s$).
} 
\label{res:results_ux_and_uy_fsi_2}
\end{figure}

\newpage
\subsection{Compliance Minimization}
In this example, we consider the compliance minimization of a standard MBB-Beam 
(Messerschmidt-B\"olkow-Blohm-Beam), see, e.g.,~\cite{BendsoeSigmund:2003}. 
In order to enforce thickness values to be either $1$ or $0$ we employ 
the, so called, SIMP-model (Solid Isotropic Material with Penalization), 
see as well,~\cite{BendsoeSigmund:2003}.

If we pick the SIMP paramter $p=1$ 
the thickness variable is allowed to take intermediate value apart from $0$ and $1$,
i.e., we consider a variable thickness sheet. The discretization is done 
using $Q_2$ elements for the displacement and discontinuous $P_0$ elements for the 
thickness.
\begin{figure}
\centering
\hspace*{\fill}
\includegraphics[width=0.44\textwidth, viewport=205 88 975 500, clip]{Pictures/MBB.png}
\hspace{\fill}
\includegraphics[width=0.44\textwidth, viewport=205 88 975 500, clip]{Pictures/MBB-top.png}
\hspace{\fill}
\caption{Thickness distribution for the MBB-beam minimum compliance problem using SIMP parameter $p=1$ (left) and $p=4$ right.} 
\label{res:mbb}
\end{figure}
A typical thickness distribution is shown in the left image in Fig.~\ref{res:mbb}.
To obtain thickness values that are, mostly, either $1$ or $0$ we need 
to pick a larger value for $p$. The distribution given in the right image in Fig.~\ref{res:mbb}
has been obtained using $p=4$.

@@

\section*{Thanks}
The \dope{} project makes use of various finite elements taken from  
the \deal{} \cite{dealnew} finite element library which has been developed
 initially by W. Bangerth, R. Hartmann, and G. Kanschat \cite{dealold}.
The authors acknowledge their past experience as well as discussions 
on modularization of algorithms
with 
the authors of the libraries 
Gascoigne, initiated by Malte Braack and  Roland Becker \cite{gascoigne}
and RoDoBo, which where initiated by 
Roland Becker, Dominik Meidner, and Boris Vexler \cite{rodobo}. 
Finally, the second author thanks Mary F. Wheeler (ICES, Austin) for the 
possibility to finish this work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\bibliographystyle{abbrvnat}
\bibliographystyle{acmsmall}
\bibliography{lit}
%

>>>>

\documentclass[prodmode,acmtoms]{acmsmall}

% Package to generate and customize Algorithm as per ACM style
\usepackage[ruled]{algorithm2e}
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

% Metadata Information
\acmVolume{?}
\acmNumber{?}
\acmArticle{??}
\acmYear{???}
\acmMonth{0}

@@

\markboth{C. Goll et al.}{DOpElib: Differential Equations and Optimization Environment}

@@

%\titlerunning{DOpElib}        % if too long for running head

\author{CHRISTIAN GOLL
\affil{University of Heidelberg}
THOMAS WICK
\affil{The University of Texas at Austin}
WINNIFRIED WOLLNER
\affil{University of Hamburg}}

\begin{abstract}
In this article, we describe the  
\textit{Differential Equations and Optimization Environment} (\dope{}).
\dope{} is a software library that provides a unified interface 
to high level algorithms such as time-stepping methods, nonlinear solvers and 
optimization routines. This structure ensures 
that, first of all, the user is only required to write those sections of code 
that are specific to 
the considered problem. Second, the exchange of parts of the used routines is possible 
with only a few lines of code to change instead of large reimplementations.
The article illustrates the design principles and various features
of \dope{} and provides some 
numerical results as demonstration for the versatility of the software.
\end{abstract}

\category{G.4}{Mathematical Software}{Optimization routines and interface to finite element software}
\category{G.1.6}{Optimization}{Optimization with partial differential equations -- Constrained optimization}
\category{G.1.8}{Partial Differential Equations}{Finite element method}
%\todocg{ Siehe http://www.acm.org/about/class/1998/, ich komme damit aber nicht zurecht.}

\terms{Algorithms, Design, Documentation}

\keywords{goal oriented interface design, numerical solution of optimization problems and partial differential equations}

\acmformat{Goll, C.; Wick, G.; Wollner, W.; DOpElib: Differential Equations and Optimization Environment.}

\begin{bottomstuff}
Author's addresses: C. Goll, Institute of Applied Mathematics, 
              University of Heidelberg,
              Im Neuenheimer Feld 293/294, 69120 Heidelberg, Germany,  christian.goll@iwr.uni-heidelberg.de;   
T. Wick, The Institute for Computational Engineering and Sciences,
The University of Texas at Austin,
Austin, Texas 78712, USA, 
              twick@ices.utexas.edu;   
Winnifried Wollner,  
              Universit\"at Hamburg,
              Department of Mathematics,
              Bundesstr. 55, 20146 Hamburg, Germany,winnifried.wollner@math.uni-hamburg.de;   
\end{bottomstuff}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exists a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \citep{dealnew}, dune \citep{dune}, 
FEniCS \citep{fenics}
or commercial solvers as for instance ANSYS (with its Fluent package),
COMSOL, or simufact forming.
Finally, a lot of work is (still) done in Matlab.
In fact, these libraries provide very useful tools 
for a huge variety of applications.

@@

Now, we describe a program designed to compute a value 
$J(u)$, where $u=(v,p)$ is the solution of the Navier-Stokes 
equations \eqref{eq:ns} on the domain $\Omega$ as seen in 
Fig.~\ref{fig:example_ns} with  $f=0$, and $J$ is the drag-force 
of the enclosed circle. We assume given Dirichlet data for the velocity 
$v$ on the boundary $\Gamma_{D}=\Gamma_{in}\cup\Gamma_{wall}\cup\Gamma_{circ}$. 
This is the Benchmark-Case 2D-1 described in \cite{TuSchae96}.
\begin{figure}[hb]
\centering
\resizebox{0.5\textwidth}{!}{\input{Pictures/example_ns}}
\caption{Flow around a cylinder with 
circle-center $C=(0.2,0.2)$ and radius $r=0.05$.}
\label{fig:example_ns}
\end{figure}

@@

\begin{figure}[h]
\centering
{\includegraphics[trim = 50mm 50mm 50mm 50mm, clip, width=0.95\textwidth]{Pictures/local_grid_NSv2.png}}
\caption{Locally adapted grid after eight refinement steps.}
\label{fig:NS_local_grid}
\end{figure}

@@

The quantities of interest are evaluations of 
$x$- and $y$ displacement at the point $A(0) = (0.6,0.2)$
and the drag and lift forces acting on the cylinder and the elastic beam:
\begin{align}
\label{drag_lift_forces}
(F_D , F_L) 
= {\int_{S_f} \sigma_f \cdot n_f \, ds + 
\int_{\Gamma_i} \sigma_s \cdot n_s \, ds},
\end{align}
where $S_{f}$ denotes the path over the cylinder in the fluid part and
$\Gamma_i$ the interface between the elastic beam and the 
fluid.
% Bilder von FSI 2 
\begin{figure}[hbt]
\centering
{\includegraphics[width=5.5cm]{Pictures/visit_fsi_2_CNn_t_2e-2_global_3_biharmonic_mesh8070_scale.png}}
{\includegraphics[width=5.5cm]{Pictures/visit_fsi_2_CNn_t_2e-2_global_3_biharmonic_x_velo8070_scale.png}}
\caption{FSI 2 test case: mesh (left) and velocity profile in vertical 
direction (right) at time $t=16.14s$.}
\label{res:fsi_2_mesh_and_x_velo}
\end{figure}

@@

\begin{figure}[hbt]
\centering
{\includegraphics[width=0.42\textwidth]{Pictures/ux_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=0.42\textwidth]{Pictures/uy_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=0.42\textwidth]{Pictures/Drag_fluid_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=0.42\textwidth]{Pictures/Lift_fluid_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
\caption{FSI 2: The deflections of the beam, $u_x(A)$ and $u_y(A)$ (in $cm$), and 
the drag $F_D$ and the lift $F_L$ evaluation (in $kg/m\,s^2$) are displayed versus
time (in $s$).
} 
\label{res:results_ux_and_uy_fsi_2}
\end{figure}

\subsection{Compliance Minimization}
In this example, we consider the compliance minimization of a standard MBB-Beam 
(Messerschmidt-B\"olkow-Blohm-Beam), see, e.g.,~\cite{BendsoeSigmund:2003}. 
In order to enforce thickness values to be either $1$ or $0$ we employ 
the, so called, SIMP-model (Solid Isotropic Material with Penalization), 
see as well,~\cite{BendsoeSigmund:2003}.

If we pick the SIMP paramter $p=1$ 
the thickness variable is allowed to take intermediate value apart from $0$ and $1$,
i.e., we consider a variable thickness sheet. The discretization is done 
using $Q_2$ elements for the displacement and discontinuous $P_0$ elements for the 
thickness.
\begin{figure}[hbt]
\centering
\hspace*{\fill}
\includegraphics[width=0.44\textwidth, viewport=205 88 975 500, clip]{Pictures/MBB.png}
\hspace{\fill}
\includegraphics[width=0.44\textwidth, viewport=205 88 975 500, clip]{Pictures/MBB-top.png}
\hspace{\fill}
\caption{Thickness distribution for the MBB-beam minimum compliance problem using SIMP parameter $p=1$ (left) and $p=4$ right.} 
\label{res:mbb}
\end{figure}
A typical thickness distribution is shown in the left image in Fig.~\ref{res:mbb}.
To obtain thickness values that are, mostly, either $1$ or $0$ we need 
to pick a larger value for $p$. The distribution given in the right image in Fig.~\ref{res:mbb}
has been obtained using $p=4$.

@@

%\section*{Thanks}
\begin{acks}
The \dope{} project makes use of various finite elements taken from  
the \deal{} \cite{dealnew} finite element library which has been developed
 initially by W. Bangerth, R. Hartmann, and G. Kanschat \cite{dealold}.
The authors acknowledge their past experience as well as discussions 
on modularization of algorithms
with 
the authors of the libraries 
Gascoigne, initiated by Malte Braack and  Roland Becker \cite{gascoigne}
and RoDoBo, which where initiated by 
Roland Becker, Dominik Meidner, and Boris Vexler \cite{rodobo}. 
Finally, the second author thanks Mary F. Wheeler (ICES, Austin) for the 
possibility to finish this work.
\end{acks}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\bibliographystyle{abbrvnat}
\bibliographystyle{acmsmall}
\bibliography{lit}
% History dates
\received{??}{??}{??}
